---
title: "Delta EAF Workflow"
format:
  html:
    toc: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{Delta EAF Workflow}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

ggplot2::theme_set(ggplot2::theme_bw(base_family = "helvetica"))
```

```{r}
#| label: setup
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
library(ggupset)
library(tidyr)
library(tibble)
library(qSIP2)
packageVersion("qSIP2")
```

# Background

## What is delta EAF?

Delta EAF quantifies the difference in isotope incorporation between two experimental treatments at the feature level. Standard EAF analysis yields a single per-feature value per comparison, with confidence intervals derived from bootstrap resampling. When comparing treatments directly — "does this feature incorporate more label in Treatment A than Treatment B?" — those point estimates alone are insufficient for inference. Delta EAF addresses this by using the same bootstrap resamples already computed during EAF calculations to build a distribution of pairwise differences, from which a confidence interval and p-value can be derived.

## Terminology

> **Note:** Throughout this vignette, a *comparison* refers to a specific pairing of labeled and unlabeled sources — one EAF calculation. A *contrast* refers to the comparison of two comparisons: which is the treatment and which is the control. A *group* is the label assigned to a comparison. The delta is always calculated as $treatment - control$, so positive values indicate higher incorporation in the treatment.

# Building a comparison list

The delta EAF functions require a named list of `qsip_data` objects — one per comparison group. Each group runs its own independent filtering and EAF calculations, so the same feature may pass in one group but not another. Here, we build four comparison groups: Normal and Drought each paired against their matched ^12^C sources, and both again paired against *all* ^12^C sources.

```{r}
q <- tribble(
  ~group,        ~unlabeled,               ~labeled,                 ~min_unlabeled_sources,
  "Normal",      "S149, S150, S151, S152", "S178, S179, S180",       3,
  "Drought",     "S161, S162, S163, S164", "S200, S201, S202, S203", 3,
  "Normal_all",  "12C",                    "S178, S179, S180",       6,
  "Drought_all", "12C",                    "S200, S201, S202, S203", 6
) |>
  run_comparison_groups(example_qsip_object,
    allow_failures = TRUE,
    seed = 41
  )
```

Because delta EAF requires a feature to have an EAF value in both groups of a contrast, features that pass filtering in one group but not another cannot be used for that contrast. `get_overlap_sizes()` counts how many features are shared across groups (@tbl-overlaps), and an upset plot gives a visual breakdown of the overlaps (@fig-upset_plot).

```{r, echo = 1}
#| label: tbl-overlaps
#| tbl-cap: "Number of features with EAF values shared across each combination of comparison groups."

overlaps = get_overlap_sizes(q)

overlaps |>
  knitr::kable()
```

```{r}
#| message: false
#| warning: false
#| label: fig-upset_plot
#| fig-cap: "Feature overlap across comparison groups, showing how many features have EAF values in each combination."
#| fig-height: 4
#| fig-width: 6
lapply(q, get_EAF_data) |>
  bind_rows(.id = "group") |>
  filter(is.na(resample)) |>
  select(group, feature_id) |>
  summarize(group = list(group), .by = feature_id) |>
  ggplot(aes(x = group)) +
    geom_bar() +
    ggupset::scale_x_upset() +
    geom_bar(fill = "#216AD0") +
    geom_text(stat='count', aes(label=after_stat(count)), vjust=-0.3, color = "gray30")
```

# Delta EAF

## Identify the contrasts

To run the delta EAF calculations, you must first define your *contrasts*. If you don't define the contrasts up front the function will do an *all-by-all* comparison, but this may not be ideal because 1) not all contrasts are meaningful, and 2) you can't determine which is the *treatment* and which is the *control*. At a high level, the *delta* is the difference in resampled EAF distributions between the treatment and the control — calculated as $treatment - control$ — so positive values indicate higher EAF in the treatment and negative values indicate higher EAF in the control. Knowing which group plays which role is critical for interpretation.

The `make_delta_EAF_contrasts()` will make a tibble of contrasts for you from your qSIP2 list object, again in an *all-by-all* format and with treatment/control pairings that may or may not be what you want (@tbl-default_contrasts).

```{r, echo = 1}
#| label: tbl-default_contrasts
#| tbl-cap: "Default contrasts generated by make_delta_EAF_contrasts(), which uses an all-by-all approach and may not reflect the intended control/treatment pairings."

contrasts = make_delta_EAF_contrasts(q)

contrasts |>
  knitr::kable()
```

For example, it is interesting to compare the drought to normal, and maybe each of those treatments against their "_all" counterpart... but perhaps the "Drought vs Normal_all" and "Drought_all vs Normal" contrasts aren't that useful. Additionally, you may not like the default contrast names which goes under the format of *control vs treatment*. To fix both of these issues, you can do an inline `filter()` and `mutate()` to get the contrasts exactly how you want it (@tbl-better_contrasts).

```{r, echo = 1}
#| label: tbl-better_contrasts
#| tbl-cap: "Filtered and renamed contrasts table. The default output can also be replaced entirely by creating a custom tibble with tribble() or reading from an Excel file."

contrasts = make_delta_EAF_contrasts(q) |>
  dplyr::filter(!contrast %in% c("Drought vs Normal_all", "Drought_all vs Normal")) |>
  mutate(contrast = paste(treatment, control, sep = " minus "))

contrasts |> knitr::kable()
```

## Run delta EAF contrasts

The `run_delta_EAF_contrasts()` function takes our qSIP2 list object (and optional contrasts) and generates a tibble with a line of output for each feature_id in each contrast. Since we already have a contrasts file, we can provide it. If we left it empty, then it would generate the same contrasts as the base `make_delta_EAF_contrasts()` function. We also give it the confidence level of 95%. 

```{r}
#| label: run_delta_EAF_contrasts
delta_EAF = run_delta_EAF_contrasts(q, 
                                    contrasts = contrasts,
                                    confidence = 0.95) 
```

The output includes two warnings — "74 contrast" and "66 `bs_pval`" — which are addressed in turn. The contrast warning indicates that 74 features could not have a delta EAF calculated for all contrasts, because they were filtered out of one or both groups used in a given contrast. With 4 contrasts, each feature would normally produce 4 rows, but features missing from any group will have fewer. @fig-contrast_warnings shows that 162 features appear in all 4 contrasts, 40 appear in only one, and 17 appear in only two.


```{r}
#| echo: false
#| label: fig-contrast_warnings
#| fig-cap: "Distribution of features by the number of contrasts in which they appear, out of a maximum of 4."
#| fig-height: 3
#| fig-width: 2
delta_EAF |>
  count(feature_id) |>
  ggplot(aes(x = n)) +
    geom_histogram(binwidth = 1, fill = "#216AD0") +
    stat_bin(
      binwidth = 1,
      geom = "text",
      color = "white",
      aes(label = after_stat(count)), # Map the 'count' computed by stat_bin to the label
      vjust = 2 # Adjust vertical position slightly above the bars
  )
```

The 17 features appearing in two contrasts each contribute two failed contrasts, giving `17+17+40 = 74`. The message is printed once per *successful* contrast — so for `ASV_167`, which had two successful and two failed contrasts, the same message appears twice. This is intentional: when viewing either successful row for that feature, the message makes clear that not all contrasts succeeded.

```{r}
#| echo: false
delta_EAF |> 
  dplyr::filter(feature_id == "ASV_167") |>
  select(feature_id, contrast_message) |>
  knitr::kable()
```

> **Note:** Warnings are only generated for features that appear in at least one contrast. A feature present in only one group — and therefore not valid in any contrast — will be absent from the output table entirely and will not generate a warning.

# Inspecting results

The results are stored in the `delta_EAF` dataframe, with one row per feature per contrast. `ASV_10` is a useful example because it reaches significance in some contrasts but not others.

## Example feature: ASV_10

```{r, echo = 1}
#| label: tbl-ASV_10
#| tbl-cap: "Delta EAF results for ASV_10 across all four contrasts."
ASV_10 = delta_EAF |>
  filter(feature_id == "ASV_10")

ASV_10 |>
  knitr::kable()
```

Two of the 4 `bs_pval` p-values (explained below) are "significant", and two are not. Interestingly, the non-significant values come from contrasts comparing the "in-treatment" unlabeled sources vs. using "all-treatment" unlabeled. So, although they did give different EAF values, they were not deemed significantly different via the delta EAF method. 

The two significant differences are between both types of Drought vs Normal comparisons, with Normal having higher EAF than Drought for this feature. The `bs_pval` calculation is explained further below, but briefly: the null hypothesis is that the true delta EAF is approximately zero, and these two contrasts provide evidence against that.

> **Note:** The output contains two p-values. `bs_pval` is a bootstrap percentile p-value derived directly from the empirical distribution of resampled delta values — it asks what proportion of the bootstrap deltas fall on each side of zero and computes a two-sided p-value from that. `pval` is a Wald z-test calculated as $2 \times \Phi(-|\Delta / \text{SD}|)$, where $\Delta$ is the observed delta EAF and $\text{SD}$ is the standard deviation of the bootstrap delta distribution. `bs_pval` is recommended for most purposes: the distribution of resampled deltas can be non-normal, particularly with few replicates, and `bs_pval` makes no parametric assumptions. `pval` is retained as an alternative for cases where a parametric test is preferred.

## A more global look

Plotting the `delta` values for each contrast together shows that most features cluster near zero, indicating little difference in incorporation between treatments (@fig-violin). But in some cases there are clear global shifts — for example, Normal_all appears to have more overall enrichment than Drought_all.

```{r}
#| fig-height: 7
#| fig-width: 9
#| label: fig-violin
#| fig-cap: "Delta EAF distributions for all features across the four contrasts, with points colored by Benjamini-Hochberg adjusted significance."
#| warning: false
#| echo: false

sig_pval = 0.05
eaf_diff = 0.05

delta_EAF |>
  mutate(bs_pval_BH = p.adjust(bs_pval)) |>
  mutate(sig = case_when(
    bs_pval_BH < sig_pval ~ TRUE,
    .default = FALSE
  )) |>
  ggplot(aes(x = contrast, y = delta)) +
    ggbeeswarm::geom_beeswarm(aes(fill = sig), pch = 21) +
    labs(y = "delta EAF", x = "Contrast", fill = paste0("Significant at p-val < ", sig_pval)) +
    geom_violin(width = 1.5, fill = NA)
```


```{r}
#| eval: false
#| echo: false
delta_EAF |>
  arrange(bs_pval) |>
  mutate(bs_BH = p.adjust(bs_pval)) |>
  mutate(sig = ifelse(bs_BH <= 0.05, "#216AD0", "firebrick")) |>
  ggplot(aes(x = bs_pval, fill = sig)) +
    geom_histogram() +
    scale_fill_identity() +
    facet_wrap(~contrast, nrow = 1)

```

### Volcano plots

One way to show this data that fits intuitively with transcriptomics results is the volcano plot. Similar to transcriptomes, the x-axis displays the difference in EAF values (as opposed to expression differences) with point to the right of the line being higher in the control, and to the left of the line being higher in the treatment. The y-axis displays the "significance", here as the negative log10 of `bs_pval`. 

```{r}
#| fig-width: 10
#| fig-height: 4
#| label: fig-volcano
#| fig-cap: "Volcano plot of delta EAF values across contrasts. Features meeting both the significance and effect size thresholds are highlighted."

sig_pval = 0.05
eaf_diff = 0.05

delta_EAF |>
  mutate(sig = case_when(
    abs(delta) >= eaf_diff & bs_pval < sig_pval ~ TRUE,
    .default = FALSE
  )) |>
    ggplot(aes(x = delta, y = -log10(bs_pval+0.001))) +
      geom_point(aes(fill = sig), pch = 21, size = 2, alpha = 0.7) +
      facet_wrap(~contrast, nrow = 1) +
      expand_limits(y = 3.1) +
      theme(legend.position = "bottom") +
      scale_fill_viridis_d() +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      labs(x = "delta EAF",
           fill = paste0("Significant at p-val < ", sig_pval, " and EAF diff > ", eaf_diff))
```

## bs_pval error message

The delta EAF output includes two message columns. The `contrast_message` column was covered above; `bs_pval_message` is explained here.

Because calculating delta requires pairing the EAF values based on their resample number, this can fail in two cases. 

The first is simply if the two groups did a different number of resamples, say 1000 in group A and 500 in group B, then obviously we can only pair up resamples 1-500 and have to skip 501-1000 for group A. 

The second case is when we use `allow_failures = TRUE` during the resampling step (see the [resampling vignette](resampling.html)). Here, although there may be 1000 resamples in both groups, there may be sporadic `NA` values filled in when the resampling failed. 

```{r}
#| label: tbl-bs_pval_message
#| tbl-cap: "Three randomly sampled rows with a bs_pval_message, showing how many resamples were skipped for each."
#| echo: false
delta_EAF |>
  filter(!is.na(bs_pval_message)) |>
  slice_sample(n = 3) |>
  select(feature_id, contrast, bs_pval, bs_pval_message) |>
  knitr::kable()
```

The `bs_pval_message` records how many resamples were skipped because a pairwise delta could not be computed.


# The delta EAF calculations

The delta EAF calculations reuse the same bootstrap distributions already computed during EAF calculations. Within each `qsip_data` object, every feature has a collection of resampled EAF values accessible via `get_EAF_data()`. The `resample` column identifies each bootstrap iteration; the observed (non-resampled) value is stored with `resample = NA` and can be excluded. Below, the 1000 resampled EAF values for `ASV_10` are retrieved across all four comparison groups.

## Resampled EAF distributions

```{r, echo = 1}
#| label: tbl-get_EAF_data
#| tbl-cap: "The first few rows of resampled EAF values for ASV_10 across all four comparison groups, ordered by resample number."
ASV_10_EAF = get_EAF_data(q) |>
  filter(feature_id == "ASV_10") |>
  filter(!is.na(resample))

ASV_10_EAF |>
  arrange(resample) |>
  head(8) |>
  knitr::kable()
```

Plotting those values on a density plot shows the overall shape of the 4 distributions. Remember, each feature will have its own distribution shape.

```{r}
#| label: fig-ASV_10_dist
#| fig-cap: Distribution of 1000 resampled EAF values for feature ASV_10.
#| fig-height: 4
#| fig-width: 4
#| echo: false
ASV_10_EAF |>
  ggplot(aes(x = EAF, color = group)) +
    geom_density() +
    scale_color_viridis_d()
```

## Delta distributions

The first step in calculating the *delta* value is constructing the resampled delta distribution for each contrast. This is done by pairing resample 1 from the treatment with resample 1 from the control, then resample 2, and so on, producing a distribution of delta EAF values with length equal to the smaller of the two distributions. Delta is always calculated as `treatment - control`.

```{r}
#| label: tbl-ASV_10_deltas
#| tbl-cap: "Pairwise delta calculations for ASV_10 in the 'Normal minus Drought' contrast, showing treatment, control, and delta EAF for each resample."
#| echo: false
ASV_10_EAF = ASV_10_EAF |>
  filter(group %in% c(contrasts$control[2], contrasts$treatment[2])) |>
  select(group, resample, EAF) |>
  pivot_wider(names_from = group, values_from = EAF) |>
  rename(control = all_of(contrasts$control[2]),
         treatment = all_of(contrasts$treatment[2])) |>
  mutate(delta = treatment - control,
         contrast = contrasts$contrast[2]) |>
  select(contrast, resample, treatment, control, delta)

ASV_10_EAF |>
  head(10) |>
  knitr::kable()
```

```{r}
#| label: fig-ASV_10_results
#| fig-cap: "Resampled EAF distributions for the treatment (Normal), control (Drought), and their pairwise delta for ASV_10."
#| echo: false
ASV_10_values = delta_EAF |> 
  filter(feature_id == "ASV_10") |> 
  filter(contrast == "Normal minus Drought")

ASV_10_EAF |>
  pivot_longer(cols = c(-contrast, -resample), names_to = "type") |>
  ggplot(aes(x = value, color = type)) +
    geom_density() +
    scale_color_viridis_d()
```

In @fig-ASV_10_results, the green line represents the delta EAF distribution, which is the yellow line minus the purple line. This delta distribution is still literally just the final EAF values from the control group subtracted from the treatment group, but the confidence intervals and `bs_pval` are derived from this distribution of delta values. The 95% CI has the value for the 2.5% and 97.5% values in the distribution.

# Conclusion

Delta EAF extends the standard EAF workflow by leveraging existing bootstrap resamples to test whether isotope incorporation differs meaningfully between experimental treatments. The `make_delta_EAF_contrasts()` and `run_delta_EAF_contrasts()` functions provide a flexible way to define and run contrasts, and the `bs_pval` and confidence intervals give a principled basis for interpretation. For studies with the necessary data, the [growth vignette](growth.html) covers an additional workflow step for estimating per-capita growth rates.
