---
title: "Delta EAF Workflow"
format:
  html:
    toc: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{Delta EAF Workflow}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

ggplot2::theme_set(ggplot2::theme_bw(base_family = "helvetica"))
```

```{r setup}
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
library(ggupset)
library(tidyr)
library(tibble)
library(qSIP2)
packageVersion("qSIP2")
```

# Background

This is a demo of the delta EAF method introduced in `qSIP2 v0.22`. The idea behind delta EAF is to answer the question "does my feature have a different level of incorporation in this treatment vs that treatment?" The tricky part about this calculation using raw EAF values is that you only get one per treatment - all replicates and any original idea of obtaining sample variance goes out the window, and you're left asking "is 4.3% different than 8.1%?", which is a difficult question.

We did, however, at one point have replicates... and what about those bootstraps?

So, let's dive in to this delta EAF calculation and see if and how it overcomes the initial limitations.

Note, through `qSIP2` the term "comparisons" is meant to imply the original idea of which labeled sources do you want to *compare* with which unlabeled sources. For delta EAF, the term *contrast* is used to say which comparison is *contrasted* with another comparison. Hopefully this terminology doesn't trip anyone up or lead to confusion. Not helping things, the idea of a *group* is really referring to a comparison. 

## Build qSIP2 list

The delta EAF function require a `qSIP2` list, so we will make that here. Like any other qSIP2 list object, each *group* will give an EAF value for any features that pass that groups independent filtering steps. This means you can use the list to compare between experimental treatments, but also see how different filtering parameters or other decision affect the outcome. Here, we will make an object for drought and normal treatments using the built in object compared just to their 12C sample pairs, or against *all* 12C samples, giving us 4 total groups.

```{r}
q <- tribble(
  ~group,        ~unlabeled,               ~labeled,                 ~min_unlabeled_sources,
  "Normal",      "S149, S150, S151, S152", "S178, S179, S180",       3,
  "Drought",     "S161, S162, S163, S164", "S200, S201, S202, S203", 3,
  "Normal_all",  "12C",                    "S178, S179, S180",       6,
  "Drought_all", "12C",                    "S200, S201, S202, S203", 6
) |>
  run_comparison_groups(example_qsip_object,
    allow_failures = T,
    seed = 41
  )
```

Importantly, the idea of a "Delta EAF" implies we are looking to quantify the differences between treatments, and so therefore for a certain feature it must have an EAF value in both of those treatments. Since each original EAF calculation is done on a different set of source material, a feature may pass the filtering in one comparison, but not the other. In @tbl-overlaps we run a check to count just how many features are even comparable between all treatments. 

```{r, echo = 1}
#| label: tbl-overlaps

overlaps = get_overlap_sizes(q)

overlaps |>
  knitr::kable()
```

Although not yet a stand alone `qSIP2` function, we can make an upset plot showing a similar idea (@fig-upset_plot).

```{r}
#| message: false
#| warning: false
#| label: fig-upset_plot
#| fig-cap: Filtered feature_ids shared between groups
#| fig-height: 4
#| fig-width: 6
lapply(q, get_EAF_data) |>
  bind_rows(.id = "group") |> 
  filter(is.na(resample)) |>
  select(group, feature_id) |>
  summarize(group = list(group), .by = feature_id) |>
  ggplot(aes(x = group)) +  
    geom_bar() +  
    ggupset::scale_x_upset() +
    geom_bar(fill = "#216AD0") +  
    geom_text(stat='count', aes(label=after_stat(count)), vjust=-0.3, color = "gray30")
```

Again, these are just some checks to see which feature_ids it is even possible to get a delta EAF for. 

# Delta EAF

## Identify the contrasts

To run the delta EAF calculations, you must first define your *contrasts*. If you don't define the contrasts up front the function will do an *all-by-all* comparison, but this may not be ideal because 1) not all contrasts are meaningful, and 2) you can't determine which is the *treatment* and which is the *control*. We'll go over the details of the delta EAF function below, but at a high level the *delta* of delta EAF is looking at the differences in some distribution of EAF values between the treatment and the control. This delta is $treatment - control$, and so positive values indicate a higher EAF in the the treatment, and negative values indicate a higher EAF in the control. In your contrasts, knowing which is the treatment and which is the control is critical for interpretation. 

The `make_delta_EAF_contrasts()` will make a tibble of contrasts for you from your qSIP2 list object, again in an *all-by-all* format and with treatment/control pairings that may or may not be what you want (@tbl-default_contrasts).

```{r, echo = 1}
#| label: tbl-default_contrasts
#| tbl-cap: The default contrasts are just best guesses and may not be relevant control/treatment combinations, or may not accurately pick which is the control

contrasts = make_delta_EAF_contrasts(q)

contrasts |>
  knitr::kable()
```

For example, it is interesting to compare the drought to normal, and maybe each of those treatments against their "_all" counterpart... but perhaps the "Drought vs Normal_all" and "Drought_all vs Normal" contrasts aren't that useful. Additionally, you may not like the default contrast names which goes under the format of *control vs treatment*. To fix both of these issues, you can do an inline `filter()` and `mutate()` to get the contrasts exactly how you want it (@tbl-better_contrasts).

```{r, echo = 1}
#| label: tbl-better_contrasts
#| tbl-cap: The contrasts table can be modified to make it more useful, or you can create one from scratch using tribble or an excel file.

contrasts = make_delta_EAF_contrasts(q) |>
  dplyr::filter(!contrast %in% c("Drought vs Normal_all", "Drought_all vs Normal")) |>
  mutate(contrast = paste(treatment, control, sep = " minus "))

contrasts |> knitr::kable()
```

## Run delta EAF contrasts

The `run_delta_EAF_contrasts()` function takes our qSIP2 list object (and optional contrasts) and generates a tibble with a line of output for each feature_id in each contrast. Since we already have a contrasts file, we can provide it. If we left it empty, then it would generate the same contrasts as the base `make_delta_EAF_contrasts()` function. We also give it the confidence level of 95%. 

```{r}
#| label: run_delta_EAF_contrasts
delta_EAF = run_delta_EAF_contrasts(q, 
                                    contrasts = contrasts,
                                    confidence = 0.95) 
```

Here, we get a warning that there were "74 contrast" and "66 `bs_pval`" warnings. I'll explain the contrast warnings now, but save the second message explanation for later down. Simply put, the first warning is telling us that 74 feature_ids weren't able to have a delta EAF calculated for all contrasts because it was filtered out of either or both of the groupings used for that contrast. So, normally there would be 4 rows per feature_id (because our contrasts table had 4 contrasts), but some feature_ids will have less. Looking at a histogram of features and their contrast count (@fig-contrast_warnings) we see that 162 are indeed found in all 4 contrasts, but 40 are found in only one contrast, and 17 are found in only 2. 


```{r}
#| echo: false
#| label: fig-contrast_warnings
#| fig-height: 3
#| fig-width: 2
delta_EAF |>
  count(feature_id) |> 
  ggplot(aes(x = n)) +
    geom_histogram(, binwidth = 1, fill = "#216AD0") +
    stat_bin(
      binwidth = 1,
      geom = "text",
      color = "white",
      aes(label = after_stat(count)), # Map the 'count' computed by stat_bin to the label
      vjust = 2 # Adjust vertical position slightly above the bars
  )
```

That "17" gets counted twice for `17+17+40 = 74` to get us to count in the warning. For now, in each *successful* contrast that message gets printed. For example, `ASV_167` had two successful contrasts, and two failed... so the same message was printed in both successful contrasts listing the failed contrasts. This is intentional friction to make it clear when you are looking at either of these rows - only some of the contrasts were successful. 

```{r}
#| echo: false
delta_EAF |> 
  dplyr::filter(feature_id == "ASV_167") |>
  select(feature_id, contrast_message) |>
  knitr::kable()
```

Note, you'll only get a warning for features that are valid in at least one contrast. For example, if there is a feature present only in the "drought_all" group, it won't be valid in any of the contrasts and therefore won't be present in the output table nor generate a warning. 

# Inspecting results

Our results are stored in the `delta_EAF` dataframe. We can pick an example feature_id to look at, so let's pick `ASV_10` because it was "significant" in some contrasts, but not others. 

## ASV_10

```{r, echo = 1}
#| label: tbl-ASV_10
ASV_10 = delta_EAF |>
  filter(feature_id == "ASV_10")

ASV_10 |>
  knitr::kable()
```

Two of the 4 `bs_pval` p-values (explained below) are "significant", and two are not. Interestingly, the non-significant values come from contrasts comparing the "in-treatment" unlabeled sources vs. using "all-treatment" unlabeled. So, although they did give different EAF values, they were not deemed significantly different via the delta EAF method. 

The two significant differences were between both types of Drought vs Normal, with Normal having a higher EAF than Drought, for this feature. I'll explain how the `bs_pval` value gets calculated later, but briefly the null hypothesis is whether the true delta EAF value is approximately zero, and there is evidence for these two contrasts that it is not zero. 

## A more global look

We can plot the `delta` values for each contrast together. In @fig-violin it is clear that the widest part of the plots for each is near zero, indicating many of the features don't have a strong different in their EAF values. But in some cases there are clear global shifts, for example normal_all seems to have more overall enrichment than drought_all.

```{r}
#| fig-height: 7
#| fig-width: 9
#| label: fig-violin
#| warning: false
#| echo: false

sig_pval = 0.05
eaf_diff = 0.05

delta_EAF |>
  mutate(bs_pval_BH = p.adjust(bs_pval)) |>
  mutate(sig = case_when(
    bs_pval_BH < sig_pval ~ TRUE,
    .default = FALSE
  )) |>
  ggplot(aes(x = contrast, y = delta)) +
    ggbeeswarm::geom_beeswarm(aes(fill = sig), pch = 21) +
    labs(y = "delta EAF", x = "Contrast", fill = paste0("Significant at p-val < ", sig_pval)) +
    geom_violin(width = 1.5, fill = NA)
```


```{r}
#| eval: false
#| echo: false
delta_EAF |>
  arrange(bs_pval) |>
  mutate(bs_BH = p.adjust(bs_pval)) |>
  mutate(sig = ifelse(bs_BH <= 0.05, "#216AD0", "firebrick")) |>
  ggplot(aes(x = bs_pval, fill = sig)) +
    geom_histogram() +
    scale_fill_identity() +
    facet_wrap(~contrast, nrow = 1)

```

### Volcano plots

One way to show this data that fits intuitively with transcriptomics results is the volcano plot. Similar to transcriptomes, the x-axis displays the difference in EAF values (as opposed to expression differences) with point to the right of the line being higher in the control, and to the left of the line being higher in the treatment. The y-axis displays the "significance", here as the negative log10 of `bs_pval`. 

```{r}
#| fig-width: 10
#| fig-height: 4
#| label: fig-volcano

sig_pval = 0.05
eaf_diff = 0.05

delta_EAF |>
  mutate(sig = case_when(
    abs(delta) >= eaf_diff & bs_pval < sig_pval ~ TRUE,
    .default = FALSE
  )) |>
    ggplot(aes(x = delta, y = -log10(bs_pval+0.001))) +
      geom_point(aes(fill = sig), pch = 21, size = 2, alpha = 0.7) +
      facet_wrap(~contrast, nrow = 1) +
      expand_limits(y = 3.1) +
      theme(legend.position = "bottom") +
      scale_fill_viridis_d() +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      labs(x = "delta EAF",
           fill = paste0("Significant at p-val < ", sig_pval, " and EAF diff > ", eaf_diff))
```

## bs_pval error message

The delta EAF workflow produces a dataframe with two columns that collect error/warning messages. The `contrast_message` was explained previously, and here the `bs_pval_message` is explained. 

Because calculating delta requires pairing the EAF values based on their resample number, this can fail in two cases. 

The first is simply if the two groups did a different number of resamples, say 1000 in group A and 500 in group B, then obviously we can only pair up resamples 1-500 and have to skip 501-1000 for group A. 

The second case is when we use `allow_failures = TRUE` during the resampling step (see the [resampling vignette](resampling.html)). Here, although there may be 1000 resamples in both groups, there may be sporadic `NA` values filled in when the resampling failed. 

```{r}
#| label: tbl-bs_pval_message
#| tbl-cap: Three random instances of a bs_pval_message warning
#| echo: false
delta_EAF |>
  filter(!is.na(bs_pval_message)) |>
  slice_sample(n = 3) |>
  select(feature_id, contrast, bs_pval, bs_pval_message) |>
  knitr::kable()
```

The warning in the `bs_pval` column is just telling us how many of the resamples were skipped because a delta couldn't be calculated. 


# The delta EAF calculations

Under the hood, the delta EAF calculations use the same resampled EAF values already used to calculate both the EAF and the EAF confidence intervals reported in `summarize_EAF_values()`. During these calculations, within an original `qSIP2` object each feature will have a collection of 1,000 resampled EAF values, together making up a distribution of resampled EAF values. This data can be accessed via the `get_EAF_data()` function. The `resample` column contains the number of that specific resampling, and the "real" data is also stored here with a `resample` value of `NA`... so we can remove that. Below, we get the 1000 resamples of our example `ASV_10` in different groups (comparisons). 

## Resampled EAF distributions

```{r, echo = 1}
#| label: tbl-get_EAF_data
#| tbl-cap: The first few lines of ASV_10_EAF, ordered by resample number
ASV_10_EAF = get_EAF_data(q) |>
  filter(feature_id == "ASV_10") |>
  filter(!is.na(resample))

ASV_10_EAF |>
  arrange(resample) |>
  head(8) |>
  knitr::kable()
```

Plotting those values on a density plot shows the overall shape of the 4 distributions. Remember, each feature_id will have it's own shapes. 

```{r}
#| label: fig-ASV_10_dist
#| fig-cap: Distribution of 1000 resampled EAF values for feature ASV_10.
#| fig-height: 4
#| fig-width: 4
#| echo: false
ASV_10_EAF |>
  ggplot(aes(x = EAF, color = group)) +
    geom_density() +
    scale_color_viridis_d()
```

## Delta distributions

The first step in calculating the singular *delta* value is getting the resampled delta values for each contrast. This is a pairwise comparison of resample 1 in both groups, then resample 2, etc, until all we have a distribution of delta EAF values the same length as the smallest distribution of that contrast. Again, delta is calculated as `treatment - control`.

```{r}
#| label: tbl-ASV_10_deltas
#| tbl-cap: The first few rows of the delta calculations for ASV_10 for just the "Normal minus Drought" contrast
#| echo: false
ASV_10_EAF = ASV_10_EAF |>
  filter(group %in% c(contrasts$control[2], contrasts$treatment[2])) |>
  select(group, resample, EAF) |>
  pivot_wider(names_from = group, values_from = EAF) |>
  rename(control = all_of(contrasts$control[2]),
         treatment = all_of(contrasts$treatment[2])) |>
  mutate(delta = treatment - control,
         contrast = contrasts$contrast[2]) |>
  select(contrast, resample, treatment, control, delta)

ASV_10_EAF |>
  head(10) |>
  knitr::kable()
```

```{r}
#| label: fig-ASV_10_results
#| echo: false
ASV_10_values = delta_EAF |> 
  filter(feature_id == "ASV_10") |> 
  filter(contrast == "Normal minus Drought")

ASV_10_EAF |>
  pivot_longer(cols = c(-contrast, -resample), names_to = "type") |>
  ggplot(aes(x = value, color = type)) +
    geom_density() +
    scale_color_viridis_d()
```

In @fig-ASV_10_results, the green line represents the delta EAF distribution, which is the yellow line minus the purple line. This delta distribution is still literally just the final EAF values from the control group subtracted from the treatment group, but the confidence intervals and `bs_pval` are derived from this distribution of delta values. The 95% CI has the value for the 2.5% and 97.5% values in the distribution.
