---
title: "Standard qSIP Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{qSIP_workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(qSIP2)
```

# Background

A complete quantitative stable isotope probing (qSIP) workflow using the `qSIP2` package starts with three input files and ends with calculated excess atom fraction (EAF) values along with a ton of intermediate data. This vignette will be a high-level walk through of the major steps with links to more specific vignettes where more detail is appropriate. 

# The Input Files

Preparing and formatting the input files is often the most tedious part of any analysis. Our goal with the rigid (and opinionated) requirements imposed by `qSIP2` will hopefully streamline the creation of these files, and automated validation checks can remove many of the common sources of error or confusion.




