---
title: "Resampling"
format:
  html:
    toc: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{Resampling}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

ggplot2::theme_set(ggplot2::theme_bw(base_family = "helvetica"))
```

```{r}
#| label: setup
#| message: false
#| warning: false

library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(qSIP2)
packageVersion("qSIP2")
```

# Background

In `qSIP2`, we use *resampling* of weighted average densities (WADs) to estimate confidence intervals both *within* replicate groups (unlabeled vs. labeled) and for the shift in mean WADs *between* these groups. This is done using a simple bootstrap procedure, where the source WADs for each `feature_id` are repeatedly sampled with replacement $n$ times.

# Resampling in R

Let's assume a WAD dataset of 4 sources labeled *A*–*D*. 

```{r}
WADs <- c(A = 1.679, B = 1.691, C = 1.692, D = 1.703)
```

To generate a bootstrap replicate, we sample these values **with replacement** until we obtain a new sample of the same size. Because sampling is done with replacement, some values may be selected more than once, while others may not be selected at all.

For example, a single bootstrap replicate might look like this:

```{r}
#| echo: false
#| label: tbl-resample_example_single
#| tbl-cap: "A single example of bootstrap draws where B is drawn twice, and C is not drawn at all."
set.seed(17)

draws <- sample(names(WADs), size = 4, replace = TRUE)

tibble::tibble(
  draw_1 = draws[1],
  draw_2 = draws[2],
  draw_3 = draws[3],
  draw_4 = draws[4]
) |> knitr::kable()
```

We can repeat this process $n$ times to generate multiple bootstrap replicates:

```{r}
#| label: tbl-resample_example
#| tbl-cap: An example of 5 resamplings of A-D
#| echo: false
n <- 5
set.seed(15)
boot_samples <- map_dfr(seq_len(n), \(i) {
  draws <- sample(names(WADs), size = 4, replace = TRUE)
  tibble(
    bootstrap = i,
    draw = 1:4,
    id = draws,
    WAD = unname(WADs[draws])
  )
})

boot_samples |>
  select(bootstrap, draw, id) |>
  mutate(draw = paste0("draw_", draw)) |>
  pivot_wider(names_from = draw, values_from = id) |>
  knitr::kable()
```

Each row represents one bootstrap replicate, and each column (`draw_1`–`draw_4`) corresponds to a draw from the original data. This format makes it clear that resampling is simply repeated random sampling with replacement, where duplicates can occur naturally.

## Under the hood

Internally, `qSIP2` reformats the resampling output to be tidy by replacing the original source names with positional column names prefixed by *type*. So if the data is the "labeled" type, the resampled columns become *labeled_1*, *labeled_2*, etc., with additional columns appended for downstream calculations.

```{r}
#| label: tbl-calculate_resampled_wads
#| tbl-cap: "Output format of one bootstrap replicate applied to the toy WAD values, showing the positional column naming used internally by qSIP2."
#| echo: false
purrr::map_df(1:n, \(i) qSIP2:::calculate_resampled_wads(i, as.data.frame(t(WADs)), "labeled")) |>
  knitr::kable()
```

> **Note:** Each resampling produces a different set of draws, but within a given resample the same draws are applied to all features. For example, if resample #1 yields `A A C D`, those same WAD values are used for every feature in that resample. This behavior can have consequences, discussed below.

# qSIP2 resampling

The `qSIP2` package has a function called `run_resampling()` that will perform the resampling procedure on a filtered `qsip_data` object. This object must first be filtered with the `run_feature_filter()` function, and we'll come back to how this filtering affects the resampling in a bit.

```{r}
q <- run_feature_filter(example_qsip_object,
  unlabeled_source_mat_ids = get_all_by_isotope(example_qsip_object, "12C"),
  labeled_source_mat_ids = c("S178", "S179", "S180"),
  min_unlabeled_sources = 3,
  min_labeled_sources = 3,
  min_unlabeled_fractions = 6,
  min_labeled_fractions = 6,
  quiet = TRUE
)

q <- run_resampling(q,
  resamples = 1000,
  with_seed = 19,
  progress = FALSE
)
```

Setting `resamples = 1000` will give 1000 resamplings for each feature. The resampling is not a deterministic procedure, and so the use of a seed is recommended using the `with_seed` argument.

The warning above indicates that 7 unlabeled features had resampling failures. We'll return to this in the [When resampling goes wrong](#when-resampling-goes-wrong) section to determine whether these failures are severe enough to be a concern.

After resampling, `get_object_summary()` confirms the updated state of the object.

```{r}
get_object_summary(q)
```

# Inspect resample results

The resampling results are stored in the `qSIP_data` object in the `@resamples` slot, but they are not necessarily intended to be worked with directly. Instead, the `qSIP_data` object has helper functions like `n_resamples()` and `resample_seed()` that will return the number of resamples that were performed and the seed that was used, respectively.

```{r}
n_resamples(q)
resample_seed(q)
```

If you want the data itself, you can access it with the `get_resample_data()` function and appropriate arguments (@tbl-get_resample_data). Note, if you set `pivot = TRUE` the dataframe can be quite large and take a while to assemble/display.

```{r, echo = 1}
#| label: tbl-get_resample_data
#| tbl-cap: The first few lines of the unlabeled resampling results. There were 8 draws because there were 8 unlabeled sources used in the filtering. 

resamples = get_resample_data(q, type = "unlabeled")

resamples |>
  head() |>
  knitr::kable()
```

## Visualizing range of mean resampled WADs

Rather than seeing all of the resampled data itself, you are often only interested in the range of the mean WAD values for each resampling iteration. You can leave the `feature_id` argument empty to see all of the features, or you can specify a single feature or a vector of features. Here, I will select 3 random feature_ids to show.

```{r}
#| fig-width: 8
#| fig-height: 3
#| label: fig-plot_feature_resamplings
#| fig-cap: Feature resampling results of 3 random features plotted with plot_feature_resamplings()

set.seed(52)
random_features <- sample(get_feature_ids(q, filtered = TRUE), 3)
plot_feature_resamplings(q,
                         feature_id = random_features,
                         intervals = "bar",
                         confidence = 0.95)
```

## Why bootstrap distributions can appear “spiky”

When the number of observations is small, bootstrap resampling can only produce a limited set of possible values. In our case, each bootstrap replicate is formed by sampling **3 values with replacement** and then taking their mean. Because the sample size is so small, many bootstrap replicates end up reusing the same values in different combinations. If one of the original values is more extreme than the others, this effect becomes more pronounced.

For example, consider three values:

```
1, 1, 10
```

When sampling 3 values with replacement, there are only a few distinct combinations (ignoring order):

- (1, 1, 1) → mean = 1  
- (1, 1, 10) → mean = 4  
- (1, 10, 10) → mean = 7  
- (10, 10, 10) → mean = 10  

Even if we generate 1,000 bootstrap replicates, the means can only take on these few values (see the points in @fig-plot_feature_resamplings). As a result, the distribution of bootstrap means is not smooth, but instead shows **discrete spikes** at these values.

This same behavior appears in the WAD data. Each feature in @tbl-wad_values has only three observed values:

```{r}
#| label: tbl-wad_values
#| tbl-cap: Actual WAD values for the three ASVs
#| echo: false
q@wads  |>
  filter(feature_id %in% random_features) |>
  left_join(get_dataframe(q, "source"), by = join_by(source_mat_id)) |>
  filter(isotope == "13C") |>
  select(feature_id, source_mat_id, WAD) |>
  pivot_wider(names_from = source_mat_id, values_from = WAD) |>
  knitr::kable()
```



Because bootstrap replicates are built from just these three values, the resulting means can only take on a limited set of possibilities. When plotted as a histogram with `points = TRUE`, this leads to the “spiky” appearance where the resampled values stack in more discrete locations rather than smoothly. Each point is mean WAD value for one of the 1000 successful resamplings. 

```{r}
#| fig-width: 8
#| fig-height: 3
#| label: fig-plot_feature_resamplings3
#| fig-cap: Feature resampling results of 3 random features with 95% CI, with points = TRUE

plot_feature_resamplings(q,
                         feature_id = random_features,
                         intervals = "bar",
                         confidence = 0.95,
                         points = TRUE)
```

In general, this effect diminishes as the number of observations increases: with more values, there are many more possible resampled combinations, and the bootstrap distribution becomes smoother. Just to further drive it home, @fig-plot_feature_resamplings3 shows some spikiness for ASV_24, even in the blue unlabeled histogram with 8 values used in the bootstrapping. If we make a plot of the WAD values used for bootstrapping, we can see an outlier in that one feature (@fig-geom_beeswarm), and that one outlier is enough to introduce spikiness in the histogram. The benefits of the bootstrapping are also apparent here because although that outlier is up near a density of 1.78 g/mL, none of the mean WAD values approach that high, unrealistic value. 

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 6
#| label: fig-geom_beeswarm
#| fig-cap: Feature resampling results of 3 random features with 95% CI
a = q@wads  |>
  filter(feature_id %in% random_features) |>
  left_join(get_dataframe(q, "source"), by = join_by(source_mat_id)) |>
  select(feature_id, source_mat_id, WAD, isotope) |>
  ggplot(aes(y = isotope, x = WAD)) +
    ggbeeswarm::geom_beeswarm(pch = 21, aes(fill = isotope), size = 2, cex = 3) +
    scale_fill_manual(values = isotope_palette) +
    facet_wrap(~feature_id, scales = "free_x") +
    theme(legend.position = "none")

b = plot_feature_occurrence(q, random_features,
                        show_wad = TRUE,
                        scale = "feature")

a / b +
  plot_layout(axes = "collect", heights = c(1,3)) +
  plot_annotation(tag_levels = "A")
```

The outlier here is from source S151 (@fig-geom_beeswarm), in which ASV_24 is found in only one fraction. This brings up an important point that although in the filtering we required a feature to be found in at least 6 fractions in at least 3 source_mat_ids, the data is still used from all sources if a feature passes. So, although ASV_24 was not found in 6 fractions in S150 and S151, and ASV_56 also didn't have enough fractions in S151, the WAD values for these cases are still used because these features overall passed the filtering criteria.

# When resampling goes wrong

The resampling procedure is a simple bootstrap procedure, and so it is not without its limitations. The most common issue is when there are more sources than your filtering requires, and you end up with WADs that contain `NA` values. 

Take ASV_72 as an example, it is found in only 1 of the labeled sources (S178) above the fraction threshold, so it was removed from the previous filtering step. But, if your filtering requirements were less strict as before (e.g. by setting `min_labeled_sources = 1`), then this feature could make it through the filtering. 

```{r}
q2 <- run_feature_filter(example_qsip_object,
  unlabeled_source_mat_ids = get_all_by_isotope(example_qsip_object, "12C"),
  labeled_source_mat_ids = c("S178", "S179", "S180"),
  min_unlabeled_sources = 3,
  min_labeled_sources = 1,
  min_unlabeled_fractions = 6,
  min_labeled_fractions = 6,
  quiet = TRUE
)
```

```{r}
#| label: tbl-asv72-wads
#| echo: false
#| tbl-cap: "WAD values for ASV_72, showing NaN for labeled sources where it did not meet the fraction threshold."
q2@wads |>
  filter(feature_id == "ASV_72") |>
  filter(source_mat_id %in% get_source_mat_ids(q, filtered = TRUE)) |>
  knitr::kable()
```

But we now get an error when running the resampling step suggesting we increase our filtering stringency. 

```{r}
#| error: true
run_resampling(q2,
  resamples = 1000,
  with_seed = 19,
  progress = FALSE
)
```

## Handling missing values during bootstrap resampling

Bootstrap resampling in `qSIP2` is performed at the level of samples (i.e., columns). For each bootstrap replicate, a set of columns is sampled with replacement, and this same set of sampled columns is applied to all features. In this way, each bootstrap replicate represents a resampled version of the original experiment.

Because the same sampled columns are used for all features, missing values (`NA`/`NaN`) are carried along in the resampling. For example, consider *ASV_72*, which has three potential source values but only one observed (non-missing) value. The remaining sources (S179 and S180) are included in the resampling but have values of `NaN`.

In this case, bootstrap replicates frequently include one or more `NaN` values. When all sampled values for a replicate are `NaN`, the mean is undefined for that feature and replicate. Rather than modifying the sampled columns on a feature-by-feature basis (e.g., by dropping columns with missing values), we retain the original resampling scheme and record the result as `NA`.

This behavior is expected: bootstrapping a vector that is largely `NaN` will often produce undefined results because the effective information for that feature is very limited. Importantly, we do not alter the sampled columns within individual features, as doing so would break the shared resampling structure across features and make bootstrap replicates no longer directly comparable.

As a result, features with many missing values (such as *ASV_72*) may contribute fewer valid bootstrap replicates. Summary statistics (e.g., confidence intervals) are therefore computed using only the valid (non-missing) bootstrap results for each feature. In practice, filtering features to require a minimum number of observed sources (e.g., via `min_labeled_sources`) can reduce the prevalence of `NaN` values entering the resampling procedure and improve stability.

## Allow failures

Although increasing the stringency can remove the error, it will also remove features from the dataset. If you want to keep the features (e.g. you didn't want *ASV_72* removed), you can set `allow_failures = TRUE` in the `run_resampling()` function. This will allow the resampling to continue, but it will also return a warning message that the resampling failed for some iterations of some features. 

```{r}
q2 <- run_resampling(q2,
  resamples = 1000,
  with_seed = 19,
  allow_failures = TRUE,
  progress = FALSE
)
```

The warning message here indicates failures in both the unlabeled and labeled resampling — 17 unlabeled and several labeled features had failures. This is similar in kind to the 7 unlabeled failures we saw in the initial `q` object, but more pronounced due to the relaxed filtering in `q2`. We can see which specific features had failures using `get_resample_counts()`, filtering for values of $n$ less than 1000 (our number of resamples).

```{r}
#| label: tbl-resample-counts
#| tbl-cap: "Features with fewer than 1000 successful resamples, indicating resampling failures."
get_resample_counts(q2) |>
  filter(labeled_resamples < 1000 | unlabeled_resamples < 1000) |>
  knitr::kable()
```

Here, we see that indeed ASV_72 was only successful in 32 and 345 of 1000 resamplings for the labeled and unlabeled, respectively. Statistically, you may conclude that 32 resamplings is still robust enough to accept the conclusion. But inspecting the plot for these 3 features shows something strange with ASV_72 and we may still choose to remove it from our analysis.

```{r}
#| label: fig-resamplings-failures
#| fig-width: 8
#| fig-height: 3
#| fig-cap: "Resampling distributions for three features including ASV_72, which had many resampling failures."
plot_feature_resamplings(q2,
                         feature_id = c("ASV_72", "ASV_155", "ASV_161"),
                         intervals = "bar",
                         points = TRUE)
```

We can further see the resampling successes for each feature with the `plot_successful_resamples()` function, and this histogram shows that most features do have 1000 successful resamplings.

```{r}
#| label: fig-successful-resamples
#| fig-width: 4
#| fig-height: 6
#| fig-cap: "Distribution of successful resampling counts per feature, showing that most features achieved all 1000 resamples."
plot_successful_resamples(q2)
```

## Using success results for further filtering

Suppose you want to overlay this success data on your final EAF plot so you can decide you have enough resampling support to trust the EAF value you obtain. This is easy to do, and you can modify the point color based on passing a threshold. We will use a success rate of 90% (i.e. 900 of 1000 resamples) as our threshold here.

Here, we can see that although most are green, ASV_72 does show up as highly enriched and with a confidence interval clear of 0. But, the red dot further flags it as suspect and warrants a deeper look.

```{r}
#| label: fig-eaf-success
#| fig-width: 8
#| fig-height: 10
#| fig-cap: "EAF values colored by resampling success rate, with ASV_72 flagged in red as failing the 90% success threshold."
EAF = run_EAF_calculations(q2)

plot_EAF_values(EAF,
                error = "ribbon",
                confidence = 0.95,
                success_ratio = 0.9,
                color_by = "success")
```

# Conclusion

Resampling is the mechanism by which `qSIP2` propagates uncertainty from WAD values through to EAF confidence intervals. Strict upfront filtering reduces the likelihood of resampling failures, but the `allow_failures = TRUE` approach offers a flexible alternative — letting the resampling step itself reveal which features have sufficient data support. In either case, `plot_successful_resamples()` and `get_resample_counts()` are useful tools for evaluating whether the resampling results are trustworthy. Once you are confident in the resampling quality, the next step is [EAF calculations](EAF.html), where the resampled WAD distributions are used to estimate isotope incorporation for each feature.
